---
title: "Patient-Level Predictors of Ebola Outcome"
subtitle: "Descriptive Exploration of the 2014–2015 Simulated Outbreak"
author: "Shri Vishalini Rajaram"
date: "May 12, 2025"
format:
  revealjs:
    theme: [simple, custom.scss]
    slide-number: true
    transition: slide
    transition-speed: fast
    center-title-slide: true
    fig-dpi: 600
    toc: false # Made custom TOC
    width: 1200
    height: 800
    footer: "BIOS:4510 - Spring 2025 | Ebola Outbreak Project"
    highlight-style: github
    progress: true
    embed-resources: true
execute:
  echo: false
  message: false
  warning: false
---

```{r setup, echo=FALSE, include=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
library(tidyverse)
library(lubridate)
library(gt)
library(forcats)
library(viridis)
library(scales)
library(webshot2)
```

```{r read_tidy_file, echo=FALSE, include=FALSE, warning=FALSE, message=FALSE}
# Reading the tidy file
ebola <- read_csv("data/ebola_combined.csv")
```

::: toc-bar

<div>
[-Background]{.active}<br>
-Patient Demographics<br>
-Clinical Milestones<br>
-Symptoms<br>
-Viral Load<br>
-Multi Comparisons<br>
-Summary<br>
</div>

::: toc-right
Why Study Ebola Outcomes?
:::
:::

### **Background**

-   Ebola virus disease (EVD) is a severe, often fatal illness caused by the Ebola virus.
-   With rapid clinical progression and high mortality, early detection and care may improve survival - yet detailed patterns remain unclear.

#### **Dataset Summary**

After pre-processing the dataset, we evaluated **`r nrow(ebola)`** patients and **`r ncol(ebola)`** predictors between **May 2014 and April 2015**.

```{r table1_Variables, echo=FALSE}
variable_table <- tribble(
  ~Category, ~Notes,
  "Patient Demographics", 
    "Includes physical traits and hospital, which reflects treatment context.",
  "Clinical Milestones", 
    "Captures key dates and durations across the illness course.",
  "Symptom Presentation", 
    "Derived from `symptoms` column using string matching; totals as `n_symptoms`.",
  "Viral Load", 
    "Lower CT = higher viral load.",
  "Clinical Outcome (Target)", 
    "Recovery status of the patient: Recover, Death, or Unknown."
)
```


```{r table1, echo=FALSE}
variable_table |>
  gt() |>
  tab_header(
    title = md("**Table 1. Variable classification used in the analysis of Ebola outbreak patient data.**")
    ) |>
  cols_label(
    Category = "Variable Category",
    Notes = "Description"
  ) |>
  fmt_markdown(columns = everything()) |>
  cols_align(align = "center", columns = everything()) |>
  tab_options(
    table.font.size = px(18),
    table.font.names = "Segoe UI, Arial",
    table.width = pct(100),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#f5f5f5",
    column_labels.border.bottom.color = "#ccc",
    column_labels.border.bottom.width = px(1),
    table.border.top.color = "#ddd",
    table.border.bottom.color = "#ddd",
    heading.title.font.size = px(24),
    row.striping.background_color = "#fbfbfb",
    row.striping.include_table_body = TRUE,
    data_row.padding = px(15),
    table.align = "left",
    row_group.padding = px(0),  
    heading.padding = px(0)
  ) |>
  opt_table_outline(style = "solid", color = "#ccc", width = px(1))
```

------------------------------------------------------------------------

::: toc-bar
<div>

[- Background]{.active}<br> - Patient Demographics<br> - Clinical Milestones<br> - Symptoms<br> - Viral Load<br> - Multi Comparisons<br> - Summary<br> - Evaluation<br>

</div>

::: toc-right
Research question
:::
:::

::: research-question
**Does the temporal changes of clinical milestones, symptom presentation, viral load, and patient demographics influence the clinical outcome of Ebola virus disease during the outbreak?**
:::

------------------------------------------------------------------------

::: toc-bar
<div>

-Background<br> [ -Patient Demographics]{.active}<br> -Clinical Milestones<br> -Symptoms<br> -Viral Load<br> -Multi Comparisons<br> -Summary<br>

</div>

::: toc-right
Distribution
:::

:::


##### **How are patient demographics and clinical characteristics distributed?**

::: {.cell layout-ncol="5"}
```{r hist-age-year, fig.width=3, fig.height=3, dpi=600, warning=FALSE}
#-----------------------Cleaned Age in Years------------------------
ggplot(ebola, aes(x = age_year)) +
  geom_histogram(binwidth = 5, fill = viridis(1), color = "white") +
  labs(title = "Distribution of Patient Age (years)", x = "Age (years)", y = "Count") +
  theme_classic(base_size = 8) + 
  annotate("text", x = -Inf, y = Inf, label = "A", hjust = 0.0, vjust = 1.0, size = 4, fontface = "bold")
```

```{r bar-age-cat, fig.width=3, fig.height=3, dpi=600, warning=FALSE}
#-----------------------Cleaned Age as Categorical Variable---------
ggplot(ebola, aes(x = age_cat, fill = age_cat)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = 0) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) + 
  scale_fill_viridis_d() +
  labs(title = "Patient Age Category Distribution", x = "Age Group", y = "Count") +
  theme_classic(base_size = 8) + theme(legend.position = "none") + 
  annotate("text", x = -Inf, y = Inf, label = "B", hjust = 0.0, vjust = 1.0, size = 4, fontface = "bold")
```

```{r bar-gender-clean, fig.width=3, fig.height=3, dpi=600, warning=FALSE}
#-----------------------Cleaned Gender-----------------------------
ggplot(ebola, aes(x = gender_clean, fill = gender_clean)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = 0) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) + 
  scale_fill_viridis_d() +
  labs(title = "Gender Distribution of Patients", x = "Gender", y = "Count") +
  theme_classic(base_size = 8) + theme(legend.position = "none") + 
  annotate("text", x = -Inf, y = Inf, label = "C", hjust = 0.0, vjust = 1.0, size = 4, fontface = "bold")
```

```{r bar-source, fig.width=3, fig.height=3, dpi=600, warning=FALSE}
#-----------------------Source-------------------------------------
ggplot(ebola, aes(x = source, fill = source)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = 0) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) + 
  scale_fill_viridis_d() +
  labs(title = "Reported Infection Source", x = "Source", y = "Count") +
  theme_classic(base_size = 8) + theme(legend.position = "none") + 
  annotate("text", x = -Inf, y = Inf, label = "D", hjust = 0.0, vjust = 1.0, size = 4, fontface = "bold")
```

```{r hist-weight-lb, fig.width=3, fig.height=3, dpi=600, warning=FALSE}
#-----------------------Weight (in Pounds)-------------------------
ggplot(ebola, aes(x = weight_lb)) +
  geom_histogram(binwidth = 10, fill = viridis(1), color = "white") +
  labs(title = "Distribution of Patient Weight (lb)", x = "Weight (lb)", y = "Count") +
  theme_classic(base_size = 8) + 
  annotate("text", x = -Inf, y = Inf, label = "E", hjust = 0.0, vjust = 1.0, size = 4, fontface = "bold")
```

```{r hist-height-ft, fig.width=3, fig.height=3, dpi=600, warning=FALSE}
#-----------------------Height (in Feet)---------------------------
ggplot(ebola, aes(x = height_ft)) +
  geom_histogram(binwidth = 0.25, fill = viridis(1), color = "white") +
  labs(title = "Distribution of Patient Height (ft)", x = "Height (ft)", y = "Count") +
  theme_classic(base_size = 8) + 
  annotate("text", x = -Inf, y = Inf, label = "F", hjust = 0.0, vjust = 1.0, size = 4, fontface = "bold")
```

```{r hist-temp-f, fig.width=3, fig.height=3, dpi=600, warning=FALSE}
#-----------------------Temperature (in Fahrenheit)-----------------
ggplot(ebola, aes(x = temp_f)) +
  geom_histogram(binwidth = 1, fill = viridis(1), color = "white") +
  labs(title = "Distribution of Body Temperature (F)", x = "Temperature (F)", y = "Count") +
  theme_classic(base_size = 8) + 
  annotate("text", x = -Inf, y = Inf, label = "G", hjust = 0.0, vjust = 1.0, size = 4, fontface = "bold")
```

```{r bar-hospital, fig.width=3., fig.height=3, dpi=600, warning=FALSE}
#-----------------------Hospital Name-------------------------------
ggplot(ebola |> filter(!is.na(name)), aes(x = fct_infreq(name), fill = name)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = 0) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) + 
  scale_fill_viridis_d(option = "D") +
  labs(title = "Hospital Representation in Dataset", x = "Hospital", y = "Patient Count") +
  theme_classic(base_size = 8) + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1)) + 
  annotate("text", x = -Inf, y = Inf, label = "H", hjust = 0.0, vjust = 1.0, size = 4, fontface = "bold")
```
:::

**A–B**: Most patients were under 20; age distribution is right-skewed; **C**: Gender distribution was nearly equal among known cases; **D**: Many infection sources were unspecified ("`other`"); **E–F**: Most weights: 100–130 lb; most heights: 4.5–5.5 ft; **G**: Typical body temperature: 100–102°F; **H**: Port Hospital had the highest patient count.

------------------------------------------------------------------------

::: toc-bar
<div>

-Background<br> [ -Patient Demographics]{.active}<br> -Clinical Milestones<br> -Symptoms<br> -Viral Load<br> -Multi Comparisons<br> -Summary<br>

</div>

::: toc-right
Relationships
:::

:::

#####  **How does demographical changes affect outcome?**

::: {.columns}

::: {.column width="50%"}


```{r age_outcome, fig.width=4, fig.height=3, dpi=600, warning=FALSE}
ebola |>
  filter(!is.na(age_cat), !is.na(outcome)) |>
  count(age_cat, outcome) |>
  group_by(age_cat) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = age_cat, y = prop, fill = outcome)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_viridis_d() +
  labs(x = "Age Group", y = "Proportion") +
  theme_classic(base_size = 7) +
  theme(legend.position = "bottom")
```

:::

::: {.column width="50%"}

```{r hosp_outcome, fig.width=4, fig.height=3, dpi=600, warning=FALSE}
ebola |>
  filter(!is.na(name), !is.na(outcome)) |>
  count(name, outcome) |>
  ggplot(aes(x = outcome, y = fct_rev(fct_infreq(name)), fill = n)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option = "D") +
  labs(x = "Outcome", y = "Hospital", fill = "Count") +
  theme_classic(base_size = 7) +
  theme(legend.position = "bottom")
```

:::

:::

**Left** Recovery and death proportions remain relatively stable from ages 0–49, but the 50–69 group shows a visible increase in recovery and slight decline in death proportion.

**Right** Port Hospital handled the highest number of patients across all outcomes, especially deaths, suggesting it served as a major treatment center. 


-------------


::: toc-bar
<div>

-Background<br> [ -Patient Demographics]{.active}<br> -Clinical Milestones<br> -Symptoms<br> -Viral Load<br> -Multi Comparisons<br> -Summary<br>

</div>

::: toc-right
Relationships
:::

:::

##### **Do overweight patients have higher risk of disease?**


```{r weight_age_outcome, fig.width=8, fig.height=3, dpi=600, warning=FALSE}
ebola |> 
  ggplot(aes(x = age_cat, y = weight_lb, fill = outcome)) +
  geom_violin(alpha = 0.6, position = position_dodge(width = 0.8)) +
  labs(x = "Age Category", y = "Weight (lb)", title = "Weight by Age Group and Outcome") +
  theme_classic(base_size = 7)
```


Weight increases with age as expected, but within each age group, clinical outcomes are not clearly separated by weight. 


-----------


::: toc-bar
<div>
-Background<br>
-Patient Demographics<br>
[-Clinical Milestones]{.active}<br>
-Symptoms<br>
-Viral Load<br>
-Multi Comparisons<br>
-Summary<br>

</div>

::: toc-right
Distribution
:::
:::

##### **Are there any patterns in the distribution of the clinical milestones?**

```{r days-to-outcome, fig.height=3, fig.width=6, dpi=600, warning=FALSE}
#-----------------------Days to Outcome---------------------------
# Prepare data in long format
days_long <- ebola |> 
  select(days_onset_to_outcome, days_infection_to_outcome, days_hospitalisation_to_outcome) |> 
  pivot_longer(
    cols = everything(),
    names_to = "event",
    values_to = "days"
  ) |>
  mutate(
    event = case_when(
      event == "days_onset_to_outcome" ~ "Symptom Onset",
      event == "days_infection_to_outcome" ~ "Infection",
      event == "days_hospitalisation_to_outcome" ~ "Hospitalisation"
    ),
    event = factor(event, levels = c("Infection", "Symptom Onset", "Hospitalisation"))
  )

# Plot
ggplot(days_long, aes(x = days)) +
  geom_histogram(binwidth = 2, fill = viridis(1), color = "white") +
  facet_wrap(
    ~ event, 
    scales = "free_y",
    labeller = labeller(event = c(
      "Infection" = "A. Infection",
      "Symptom Onset" = "B. Symptom Onset",
      "Hospitalisation" = "C. Hospitalisation"
    ))
  ) +
  labs(
    x = "Days to Outcome", y = "Patient Count"
  ) +
  theme_classic(base_size = 10)
```

**Panels A–C** show histograms of days between clinical events (A: infection, B: symptom onset, C: hospitalisation) and outcome. 

----------------

::: toc-bar
<div>
-Background<br>
-Patient Demographics<br>
[-Clinical Milestones]{.active}<br>
-Symptoms<br>
-Viral Load<br>
-Multi Comparisons<br>
-Summary<br>

</div>

::: toc-right
Distribution
:::
:::

##### **When was the peak time of the epidemic during 2014-2015 outbreak?**

```{r total-dates, fig.width=4, fig.height=2, dpi=600, warning=FALSE}
#-----------------------Time Trend of events (Month)-------------------------
# Count events per month
event_counts <- bind_rows(
  ebola |> filter(!is.na(month_onset)) |> count(month = month_onset) |> mutate(event = "Symptom Onset"),
  ebola |> filter(!is.na(month_infection)) |> count(month = month_infection) |> mutate(event = "Infection"),
  ebola |> filter(!is.na(month_hospitalisation)) |> count(month = month_hospitalisation) |> mutate(event = "Hospitalisation")
)

# Plot monthly counts
ggplot(event_counts, aes(x = month, y = n, color = event)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 0.7) +
  scale_color_viridis_d() +
  labs(
    x = "Month", y = "Number of Cases", color = "Event"
  ) +
  theme_classic(base_size = 7)
```


Monthly counts of cases for infection, symptom onset, and hospitalisation events. Negative values in A–C reflect likely data entry errors or same-day events.

------------------------------------------------------------------------

::: toc-bar
<div>
-Background<br>
-Patient Demographics<br>
[-Clinical Milestones]{.active}<br>
-Symptoms<br>
-Viral Load<br>
-Multi Comparisons<br>
-Summary<br>

</div>

::: toc-right
Relationships
:::
:::

##### **How did the hospitals adapt to the situation? Were there delays in treatment?**

::: {.columns}

::: {.column width="70%"}

```{r system_adaptation, fig.width=8, fig.height=6, dpi=600, warning=FALSE}
ebola |>
  filter(!is.na(name), !is.na(month_hospitalisation), !is.na(days_hospitalisation_to_outcome)) |>
  mutate(
    month_label = fct_inorder(format(ymd(month_hospitalisation), "%b %Y"))
  ) |>
  group_by(name, month_label, outcome) |>
  summarise(
    median_days = median(days_hospitalisation_to_outcome, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = month_label, y = median_days, group = outcome, color = outcome)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1) +
  facet_wrap(~ name, ncol = 2) +
  scale_color_viridis_d() +
  labs(
    title = "Median Hospitalisation-to-Outcome Time by Month, Outcome, and Hospital",
    x = "Month", y = "Median Days", color = "Outcome"
  ) +
  theme_classic(base_size = 8) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
:::

::: {.column width="30%"}

- `Military` and `Port Hospitals` show **stable and stratified patterns**, indicating organized clinical pathways by outcome.
- `St. Mark’s` and “`Other`” facilities exhibit **greater month-to-month variation**, suggesting either fewer cases or evolving protocols.
- There is **no strong evidence** of systemic delay over time - rather, median durations remained stable or improved.

:::

:::

------------

::: toc-bar
<div>
-Background<br>
-Patient Demographics<br>
[-Clinical Milestones]{.active}<br>
-Symptoms<br>
-Viral Load<br>
-Multi Comparisons<br>
-Summary<br>

</div>

::: toc-right
Relationships
:::
:::


##### **Did clinical milestones vary among different age groups in turn affecting outcome?**

::: {.columns}

::: {.column width=70%}

```{r age_milestone_outcome, fig.width=8, fig.height=6, dpi=600, warning=FALSE}
# Prepare data
timing_age <- ebola |>
  select(age_cat, outcome, starts_with("days_")) |>
  pivot_longer(cols = starts_with("days_"), names_to = "event", values_to = "days") |>
  mutate(
    event = recode(event,
      "days_infection_to_outcome" = "Infection",
      "days_onset_to_outcome" = "Symptom Onset",
      "days_hospitalisation_to_outcome" = "Hospitalisation"
    )
  )

# Plot
ggplot(timing_age, aes(x = age_cat, y = days, fill = outcome)) +
  geom_boxplot(outlier.size = 0.5) +
  facet_wrap(~ event) +
  scale_fill_viridis_d() +
  labs(x = "Age Category", y = "Days to Outcome") +
  coord_flip() + 
  theme_classic(base_size = 10)
```

:::

::: {.column width=30%}

- **Shorter time-to-outcome for deaths** across all ages likely reflects severe presentations or delayed hospitalisation.
- **Longer recovery durations** in older patients suggest sustained treatment efforts.
- Variation in milestone intervals by age implies that **age modified the timing of clinical trajectory**, thereby influencing the likelihood of recovery or death.

:::

:::

---------------

::: toc-bar
<div>
-Background<br>
-Patient Demographics<br>
[-Clinical Milestones]{.active}<br>
-Symptoms<br>
-Viral Load<br>
-Multi Comparisons<br>
-Summary<br>

</div>

::: toc-right
Relationships
:::
:::



```{r gt_delay_pattern, warning=FALSE}
ebola |>
  filter(!is.na(name), !is.na(outcome)) |>
  group_by(name, outcome) |>
  summarise(
    Infection = paste0(round(mean(days_infection_to_outcome, na.rm = TRUE), 1), 
                       " ± ", round(sd(days_infection_to_outcome, na.rm = TRUE), 1)),
    Onset = paste0(round(mean(days_onset_to_outcome, na.rm = TRUE), 1), 
                   " ± ", round(sd(days_onset_to_outcome, na.rm = TRUE), 1)),
    Hospitalisation = paste0(round(mean(days_hospitalisation_to_outcome, na.rm = TRUE), 1), 
                             " ± ", round(sd(days_hospitalisation_to_outcome, na.rm = TRUE), 1)),
    .groups = "drop"
  ) |>
  arrange(desc(name)) |>
  gt(groupname_col = "name") |>
  tab_header(
    title = md("**Average Time from Clinical Events to Outcome by Hospital and Outcome**")
  ) |>
  cols_label(
    name = "",
    outcome = "",
    Infection = "Infection to Outcome",
    Onset = "Onset to Outcome",
    Hospitalisation = "Hospitalisation to Outcome"
  ) |>
  tab_options(
    row_group.as_column = FALSE,
    row_group.font.weight = "bold",
    row_group.background.color = "#f5f5f5",
    table.font.size = px(14)
  )
```

------------------


::: toc-bar
<div>

-Background<br> -Patient Demographics<br> -Clinical Milestones<br> [-Symptoms]{.active}<br> -Viral Load<br> -Multi Comparisons<br> -Summary<br>

</div>

::: toc-right
Distribution
:::
:::

##### **Which was the highest recorded symptom among the patients?**


```{r symptom-indicator-bars, fig.width=4, fig.height=2, dpi=600, warning=FALSE}
#-----------------------Symptom Indicators-------------------------
symptom_vars <- c("symptom_fever", "symptom_cough", "symptom_vomit", "symptom_chills", "symptom_aches", "symptom_none")

indicator_counts <- ebola |> 
  select(all_of(symptom_vars)) |> 
  pivot_longer(cols = everything(), names_to = "symptom", values_to = "present") |> 
  filter(!is.na(present)) |> 
  count(symptom, present)

symptom_labels <- c(
  symptom_aches = "A. Aches",
  symptom_chills = "B. Chills",
  symptom_cough = "C. Cough",
  symptom_fever = "D. Fever",
  symptom_none = "E. No Symptoms",
  symptom_vomit = "F. Vomit"
)

ggplot(indicator_counts, aes(x = as.factor(present), y = n, fill = as.factor(present))) +
  geom_col(width = 0.2) +
  geom_text(aes(label = n), vjust = 0, size = 0.8) +
  facet_wrap(~ symptom, scales = "free_y", labeller = as_labeller(symptom_labels)) +
  scale_fill_viridis_d() +
  labs(x = "Reported", y = "Count") +
  theme_classic(base_size = 9) +
  theme(legend.position = "none",
        strip.text = element_text(size = 4),
        axis.text = element_text(size = 3),
        axis.title = element_text(size = 4))
```

**Panels A–F** show binary indicator counts for individual symptoms: fever, cough, vomiting, chills, aches, and absence of symptoms (“no symptoms”).

-------------------------

::: toc-bar
<div>

-Background<br> -Patient Demographics<br> -Clinical Milestones<br> [-Symptoms]{.active}<br> [-Viral Load]{.active}<br> -Multi Comparisons<br> -Summary<br>

</div>

::: toc-right
Distribution
:::
:::


::: {.cell layout-ncol=2}

##### **How many symptoms were commonly reported?**

```{r hist-n-symptoms, fig.width=1, fig.height=1, dpi=600, warning=FALSE}
#-----------------------Number of Symptoms-------------------------
ggplot(ebola, aes(x = n_symptoms)) +
  geom_bar(fill = viridis(1)) +
  labs(x = "Symptom Count", y = "Count") +
  theme_classic(base_size = 4)
```

##### **Where there any left skewness with the viral load depicting higher disease?**

```{r hist-ct-blood, fig.height=1, fig.width=1, dpi=600, warning=FALSE, fig.align='center'}
#-----------------------CT_Blood (PCR)-------------------------
ggplot(ebola, aes(x = ct_blood)) +
  geom_histogram(binwidth = 1, fill = viridis(1), color = "white") +
  labs(x = "CT Blood", y = "Count") +
  theme_classic(base_size = 4)
```

:::

------------------------------------------------------------------------

::: toc-bar
<div>

-Background<br> -Patient Demographics<br> -Clinical Milestones<br> -Symptoms<br> -Viral Load<br> [-Multi Comparisons]{.active}<br> -Summary<br>

</div>

::: toc-right
Distribution
:::
:::

##### **Were there any combination of symptoms prevalent in particular outcome?**

::: {.columns}

::: {.column width=70%}


```{r symptoms_combo_outcome, fig.height=4, fig.width=6, dpi=600, warning=FALSE}
ebola_symptoms_combo <- ebola |>
  mutate(
    symptom_combo = case_when(
      symptom_fever == 1 & symptom_cough == 1 & symptom_vomit == 1 ~ "Fever+Cough+Vomit",
      symptom_fever == 1 & symptom_vomit == 1 ~ "Fever+Vomit",
      symptom_fever == 1 & symptom_cough == 1 ~ "Fever+Cough",
      symptom_vomit == 1 & symptom_cough == 1 ~ "Vomit+Cough",
      TRUE ~ "Other"
    ))

ebola_symptoms_combo |> 
  count(symptom_combo, outcome) |>
  ggplot(aes(x = symptom_combo, y = n, fill = outcome)) +
  geom_col(position = position_dodge(width = 0.8)) +
  geom_text(aes(label = n), 
            position = position_dodge(width = 0.8), 
            vjust = -0.3, size = 3) +
  scale_fill_viridis_d() +
  labs(title = "Outcome by Symptom Combination", x = "Symptom Combo", y = "Count") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

:::

:::{.column width=30%}

Most patients with **`Fever+Cough`** or **`Fever+Cough+Vomit`** had a known outcome, with Death being the most frequent outcome in both combinations. 

:::

:::

----------

::: toc-bar
<div>

-Background<br> -Patient Demographics<br> -Clinical Milestones<br> -Symptoms<br> -Viral Load<br> [-Multi Comparisons]{.active}<br> -Summary<br>

</div>

::: toc-right
Distribution
:::
:::

##### **Does the combination of reported symptoms and age category influence the clinical outcome?**

::: {.columns}

::: {.column width=70%}


```{r symptoms_age_outcome, fig.height=4, fig.width=6, dpi=600, warning=FALSE}
ebola_symptoms_combo |>
  filter(symptom_combo != "Other", !is.na(age_cat), !is.na(outcome)) |>
  count(symptom_combo, age_cat, outcome) |>
  group_by(symptom_combo, age_cat) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = age_cat, y = prop, fill = outcome)) +
  geom_col(position = "fill") +
  facet_wrap(~ symptom_combo) +
  scale_fill_viridis_d() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Outcome Distribution by Symptom Combination and Age Category",
    x = "Age Category", y = "Proportion"
  ) +
  theme_classic(base_size = 8) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

:::

::: {.column width=30%}

- For **`Fever+Cough`**, younger age groups (`<30`) consistently show higher recovery proportions, while older age groups have slightly increased death rates.

- In **`Fever+Vomit`** and **`Vomit+Cough`**, recovery rates drop noticeably in middle-aged and older adults, with a corresponding rise in deaths, especially in the `50–69` group.

:::

:::


-------------

::: toc-bar
<div>

-Background<br> -Patient Demographics<br> -Clinical Milestones<br> -Symptoms<br> -Viral Load<br> [-Multi Comparisons]{.active}<br> -Summary<br>

</div>

::: toc-right
Risk profile
:::
:::

##### **Can we find a risk profile for the children?**


```{r risk_profile, dpi=600, warning=FALSE} 
# ---- Age 0–9 group summaries ----
ebola_0_9 <- ebola |> filter(age_cat == "0-9", outcome %in% c("Death", "Recover"))
combo_0_9 <- ebola_symptoms_combo |> filter(age_cat == "0-9", outcome %in% c("Death", "Recover"))

top_combo_0_9 <- combo_0_9 |>
  filter(symptom_combo != "Other") |>
  count(outcome, symptom_combo) |>
  group_by(outcome) |>
  mutate(pct = n / sum(n)) |>
  slice_max(n, n = 1, with_ties = FALSE) |>
  mutate(`Top Symptom Combo` = paste0(symptom_combo, " (", round(pct * 100, 1), "%)")) |>
  select(outcome, `Top Symptom Combo`)

summary_0_9 <- ebola_0_9 |>
  group_by(outcome) |>
  summarise(
    `CT` = round(mean(ct_blood < 21, na.rm = TRUE) * 100, 1),
    `Weight` = round(mean(weight_lb > 80, na.rm = TRUE) * 100, 1),
    `Infection to Outcome` = paste0(median(days_infection_to_outcome, na.rm = TRUE),
                                    " (", quantile(days_infection_to_outcome, 0.25, na.rm = TRUE),
                                    "–", quantile(days_infection_to_outcome, 0.75, na.rm = TRUE), ")"
                                    ),
    `Onset to Outcome` = paste0(median(days_onset_to_outcome, na.rm = TRUE),
                                " (", quantile(days_onset_to_outcome, 0.25, na.rm = TRUE),
                                "–", quantile(days_onset_to_outcome, 0.75, na.rm = TRUE), ")"
                                ),
    `Hospitalisation to Outcome` = paste0(median(days_hospitalisation_to_outcome, na.rm = TRUE),
                                          " (", quantile(days_hospitalisation_to_outcome, 0.25, na.rm = TRUE),
                                          "–", quantile(days_hospitalisation_to_outcome, 0.75, na.rm = TRUE), ")"
                                          ),
    `Top Hospital` = name |> na.omit() |> fct_infreq() |> as.character() |> first(),
    .groups = "drop"
  )

final_0_9 <- left_join(top_combo_0_9, summary_0_9, by = "outcome") |> mutate(age_group = "0-9")

# ---- Age 10–19 group summaries ----
ebola_10_19 <- ebola |> filter(age_cat == "10-19", outcome %in% c("Death", "Recover"))
combo_10_19 <- ebola_symptoms_combo |> filter(age_cat == "10-19", outcome %in% c("Death", "Recover"))

top_combo_10_19 <- combo_10_19 |>
  filter(symptom_combo != "Other") |>
  count(outcome, symptom_combo) |>
  group_by(outcome) |>
  mutate(pct = n / sum(n)) |>
  slice_max(n, n = 1, with_ties = FALSE) |>
  mutate(`Top Symptom Combo` = paste0(symptom_combo, " (", round(pct * 100, 1), "%)")) |>
  select(outcome, `Top Symptom Combo`)

summary_10_19 <- ebola_10_19 |>
  group_by(outcome) |>
  summarise(
    `CT` = round(mean(ct_blood < 21, na.rm = TRUE) * 100, 1),
    `Weight` = round(mean(weight_lb > 80, na.rm = TRUE) * 100, 1),
    `Infection to Outcome` = paste0(median(days_infection_to_outcome, na.rm = TRUE),
                                    " (", quantile(days_infection_to_outcome, 0.25, na.rm = TRUE),
                                    "–", quantile(days_infection_to_outcome, 0.75, na.rm = TRUE), ")"
                                    ),
    `Onset to Outcome` = paste0(median(days_onset_to_outcome, na.rm = TRUE),
                                " (", quantile(days_onset_to_outcome, 0.25, na.rm = TRUE),
                                "–", quantile(days_onset_to_outcome, 0.75, na.rm = TRUE), ")"
                                ),
    `Hospitalisation to Outcome` = paste0(median(days_hospitalisation_to_outcome, na.rm = TRUE),
                                          " (", quantile(days_hospitalisation_to_outcome, 0.25, na.rm = TRUE),
                                          "–", quantile(days_hospitalisation_to_outcome, 0.75, na.rm = TRUE), ")"
                                          ),
    `Top Hospital` = name |> na.omit() |> fct_infreq() |> as.character() |> first(),
    .groups = "drop"
  )

final_10_19 <- left_join(top_combo_10_19, summary_10_19, by = "outcome") |> mutate(age_group = "10-19")

# ---- Combine and reshape ----
combined <- bind_rows(final_0_9, final_10_19) |>
  mutate(across(c(`CT`, `Weight`), as.character)) |> 
  pivot_longer(cols = -c(outcome, age_group), names_to = "Feature", values_to = "Value") |>
  unite("age_outcome", age_group, outcome) |>
  pivot_wider(names_from = age_outcome, values_from = Value)


# ---- Create GT Table ----
gt(combined, rowname_col = "Feature") |>       
  tab_spanner(label = "Age 0–9", columns = c(`0-9_Death`, `0-9_Recover`)) |>
  tab_spanner(label = "Age 10–19", columns = c(`10-19_Death`, `10-19_Recover`)) |>
  cols_label(
    `0-9_Death` = "Death",
    `0-9_Recover` = "Recover",
    `10-19_Death` = "Death",
    `10-19_Recover` = "Recover"
  ) |>
  fmt_markdown(columns = everything()) |>
  tab_style(
    style = list(
      cell_text(weight = "bold"),
      cell_borders(sides = "all", weight = px(4), color = "#DDDDDD")
    ),
    locations = cells_stub()
  ) |>
  tab_style(
    style = cell_fill(color = "#f0f4f8"),
    locations = cells_body(columns = everything())
  ) |>
  tab_style(
    style = list(cell_text(weight = "bold")),
    locations = cells_column_spanners()
  ) |>
  tab_footnote(
    footnote = "CT < 21 was the threshold and it indicates higher viral load across both age groups.",
    locations = cells_stub(rows = "CT")
  ) |>
  tab_footnote(
    footnote = "Weight thresholds: >80 lb for Age 0–9; >120 lb for Age 10–19. (Source: CDC, NCHS, 2024)",
    locations = cells_stub(rows = "Weight")
  ) |>
  tab_footnote(
    footnote = "Median (IQR) reported for timing intervals in days.",
    locations = cells_stub(rows = c(
      "Infection to Outcome",
      "Onset to Outcome",
      "Hospitalisation to Outcome"
    ))
  ) |>
  opt_table_outline() |>
  opt_row_striping() |>
  opt_align_table_header(align = "left") |>
  tab_options(
    table.font.size = 18,
    table.font.names = "Segoe UI, Arial",
    table.width = pct(100),
    data_row.padding = px(10),
    heading.title.font.size = 24,
    column_labels.font.size = 20,
    table.border.top.width = px(4),
    table.border.bottom.width = px(4)
  )
```


----------------

::: toc-bar
<div>

-Background<br> -Patient Demographics<br> -Clinical Milestones<br> -Symptoms<br> -Viral Load<br> -Multi Comparisons<br> [-Summary]{.active}<br>

</div>

::: toc-right
Summary
:::
:::

##### **What did we find?**

:::: {.columns style="font-size: 1.5em;"}
::: {.column width="100%"}

- **Age matters:** Younger patients had more deaths; older adults showed higher recovery.  
- **Symptoms signal severity:** Fever + Cough combos were common in fatal cases.  
- **Weight $\neq$ Risk:** Heavier patients weren’t more likely to die after adjusting for age.  
- **Faster deaths, slower recoveries:** Time-to-outcome was shortest for deaths across all age groups.  
- **Hospitals adapted over time:** Port and Military hospitals showed stable care timelines.  

<div style="text-align:center; font-size:4em; color:#002366; margin-top:1em;">
  <strong>THANK YOU</strong>
</div>

:::
::::






